Using Fundamental Measure Theory to Treat the Correlation Function of the Inhomogeneous Hard-Sphere Fluid

### 2012-08-15 papers/contact/paper.tex

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.  Since the hard-sphere radius enters the FMT free energy
by means of the fundamental measures, this requires that we adjust the
convolutions to have a diameter $\sigma$ which depends on the location
of the sphere:
\begin{align}
  n_3^A(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-\sigma(\rr')/2)d\rr' \label{eq:n3A}
\end{align}
Once we have done this---and similarly for the other fundamental
measures---we derive a formula for the correlation function
$g_\sigma^A(\rr)$ by finding the pressure on the surface of any hard
spheres located at position $\rr$, and using the contact-density
theorem to extract the contact density, from which we can establish
the averaged correlation function at contact.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n(\rr) 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^A}\\
  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
The functional derivative $2 \frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$
in the above formula must be computed with the asymmetric formulation
for the fundamental measures.  In practice, we use the approximate
White Bear functional to find an approximate expression for
$g_\sigma^A$, which is given in Appendix~\ref{appendix:g-A}.  This
expression will involve convolutions of derivatives of the free
energy, making this formulation computationally more expensive than
the free energy itself.

In addition to the different densities in the denominators of Equations 
\ref{eq:g-A-exact-a} and \ref{eq:g-S-exact-a} there is also a difference 
between the treatment of the A and S cases in the derivatives 
$2 \frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$ in Equations \ref{eq:g-A-exact} 
and \ref{eq:g-S-exact}. The White Bear approximations for $A_{HS}$ 
that we use involve a number of weighted densities and for the derivatives 
of these we use two different conceptual approaches.  In the assymetric case, 
a change in $\sigma(\mathbf{r})/2$ will refer to a change in the radius of a sphere 
that is located at the position $\mathbf{r}$.  The weighted densities that are 
affected by such a change will be those that are located a distance $\sigma/2$ away 
from such a sphere.  We compute the assymetric derivatives In Appendix~\ref{appendix:g-A}.

>>>>

When we consider an inhomogeneous system, we may prefer to ask for
\emph{which} sphere we are interested in computing the contact
density.  One way to answer this is to specify a sphere at a fixed
position $\mathbf{r}$.  By making the hard-sphere radius dependent on
position, we can compute the pressure on the hard spheres located at
$\mathbf{r}$.  Since the hard-sphere radius enters the FMT free energy
by means of the fundamental measures, this requires that we adjust the
convolutions to have a diameter $\sigma$ which depends on the location
of the sphere:
\begin{align}
  n_3^A(\rr) &= \int n(\rr')\Theta(|\rr-\rr'|-\sigma(\rr')/2)d\rr' \label{eq:n3A}
\end{align}
Once we have done this---and similarly for the other fundamental
measures---we derive a formula for the correlation function
$g_\sigma^A(\rr)$ by finding the pressure on the surface of any hard
spheres located at position $\rr$, and using the contact-density
theorem to extract the contact density, from which we can establish
the averaged correlation function at contact.
\begin{align}
  p_{HS}(\mathbf{r}) &= \frac{1}{n(\rr) 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:p_{HS}^A}\\
  \ncontact(\rr) &= \frac{1}{n(\rr) k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \\
  g_\sigma^A(\rr) &= \frac{\ncontact(\rr)}{n_A(\rr)} \\
  &= \frac{1}{n(\rr) n_A(\rr)}\frac{1}{ k_BT 2\pi \sigma^2} \frac{\delta
    A_{HS}}{\delta \sigma(\mathbf{r})} \label{eq:g-A-exact}
\end{align}
The functional derivative $\frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$
in the above formula must be computed with the asymmetric formulation
for the fundamental measures.  In practice, we use the approximate
White Bear functional to find an approximate expression for
$g_\sigma^A$, which is given in Appendix~\ref{appendix:g-A}.  This
expression will involve convolutions of derivatives of the free
energy, making this formulation computationally more expensive than
the free energy itself.

In addition to the different densities in the denominators of Equations 
\ref{eq:g-A-exact-a} and \ref{eq:g-S-exact-a} there is also a difference 
between the treatment of the A and S cases in the derivatives 
$\frac{\delta A_{HS}}{\delta \sigma(\mathbf{r})}$ in Equations \ref{eq:g-A-exact} 
and \ref{eq:g-S-exact}. The White Bear approximations for $A_{HS}$ 
that we use involve a number of weighted densities and for the derivatives 
of these we use two different conceptual approaches.  In the assymetric case, 
a change in $\sigma(\mathbf{r})/2$ will refer to a change in the radius of a sphere 
that is located at the position $\mathbf{r}$.  The weighted densities that are 
affected by such a change will be those that are located a distance $\sigma/2$ away 
from such a sphere.  We compute the assymetric derivatives In Appendix~\ref{appendix:g-A}.
